---
title: "Interactive Data Visualization"
author: "<br> Chester Ismay (cismay@reed.edu) <br>"
date: "Paideia 2k16 <br><br> Slides available at <http://rpubs.com/cismay/paideia_2k16_idv> <br> Supplementary HTML file at <http://rpubs.com/cismay/paideia_2k16_idv_sup>"
output: 
  html_document:
    fig_width: 9.5
    fig_height: 5
    fig.align: center
    keep_md: true
---

When we think about data visualization, we often think of a static plot with some colors and points.  New tools (spearheaded by Hans Rosling's Gapminder project) are constantly being developed to allow us to interact dynamically with data visualizations.  I'll discuss a variety of different tools that attempt to make data come to life!

All too often statistics is thought of as a boring subject with boring plots.  New software packages and tools are being developed to better understand the relationships between variables.  I'll demonstrate a lot of these different tools and packages using the R computing language.  We'll go through a variety of different examples from multiple fields to better understand anomalies and trends in data.

```{r setup, include=FALSE}
pkg <- c("dplyr", "ggplot2", "knitr", "readr", 
         "xts", "maps", "googleVis", "DT", "rmarkdown")

new.pkg <- pkg[!(pkg %in% installed.packages())]

if (length(new.pkg)) {
  install.packages(new.pkg, repos = "http://cran.rstudio.com")
}

if(!require(revealjs)){
  devtools::install_github("jjallaire/revealjs", ref = "a4854c017eac44d969a216103551c21d66329a74")
}

if(!require(plotly)){
  devtools::install_github("ropensci/plotly")
}

if(!require(pnwflights14)){
  devtools::install_github("ismayc/pnwflights14")
}

if(!require(dygraphs)){
  devtools::install_github("rstudio/dygraphs", ref = "778acdaeb91b754412d928ea824632bceae3078b")
}

library(dplyr)
library(ggplot2)
library(revealjs)
library(knitr)
library(readr)
library(plotly)
library(dygraphs)
library(pnwflights14)
library(xts)
library(maps)
library(googleVis)
library(DT)

options(width = 100, scipen = 99)
```

```{r setup2, eval=FALSE}
pkg <- c("dplyr", "ggplot2", "knitr", "readr", 
         "xts", "maps", "googleVis", "DT", "rmarkdown")

new.pkg <- pkg[!(pkg %in% installed.packages())]

if (length(new.pkg)) {
  install.packages(new.pkg, repos = "http://cran.rstudio.com")
}

if(!require(revealjs)){
  devtools::install_github("jjallaire/revealjs", ref = "a4854c017eac44d969a216103551c21d66329a74")
}

if(!require(plotly)){
  devtools::install_github("ropensci/plotly")
}

if(!require(pnwflights14)){
  devtools::install_github("ismayc/pnwflights14")
}

if(!require(dygraphs)){
  devtools::install_github("rstudio/dygraphs", ref = "778acdaeb91b754412d928ea824632bceae3078b")
}

library(dplyr)
library(ggplot2)
library(revealjs)
library(knitr)
library(readr)
library(plotly)
library(dygraphs)
library(pnwflights14)
library(xts)
library(maps)
library(googleVis)
library(DT)

options(width = 100, scipen = 99)
```

### The __Iris__ flower data set

- Introduced by Ronald Fisher in 1936

- The data set consists of 50 samples from each of three species of Iris (_Iris setosa_, _Iris virginica_, and _Iris versicolor_). 

- Four features were measured from each sample: the length and the width of the sepals and petals, in centimetres. Based on the combination of these four features, Fisher developed a model to distinguish the species from each other.

Source: [Wikipedia](https://en.wikipedia.org/wiki/Iris_flower_data_set)

---

# Scatterplots

---

## Traditional (boring) plot

```{r}
with(iris, plot(x = Petal.Width, y = Sepal.Length))
```

---

## Prettier (not quite as boring) plot

```{r}
qplot(Petal.Width, Sepal.Length, data = iris)
```

---

## Interactive plot using `plotly`

```{r, plotly=TRUE}
ggiris <- qplot(Petal.Width, Sepal.Length, data = iris)
ggplotly(ggiris)
```

---

## Prettier interactive plot using `plotly`

```{r, plotly=TRUE}
ggiris_colored <- qplot(Petal.Width, Sepal.Length, data = iris, 
  color = Species)
ggplotly(ggiris_colored)
```


---

## Another interactive plot

```{r, plotly=TRUE}
iris %>% plot_ly(x = Petal.Width, y = Sepal.Length,
  type = "scatter", color = Species, mode = "markers")
```

---

# Scatterplots (Part Deux)

---

### Reed College majors VS Total Faculty FTE by department

- Based off [analysis](http://www.reed.edu/data-at-reed/resources/R/scatter_plots.html) done by Rich Majerus in 2014 using the `googleVis` package

- Data does not include 143 interdisciplinary majors and 9 undecided majors.  

- Majors like Bio/Chem are split between the two departments 

- General Lit/Lit majors are included with English 

- Dance majors and faculty are included with Theatre

```{r include=FALSE}
Departments = c('Art', 'Music', 'Theatre', 'Anthropology', 'Economics',
                'History', 'Political Science', 'Sociology', 'Chinese',  'Classics',
                'English', 'French', 'German', 'Russian', 'Spanish',
                'Biology', 'Chemistry', 'Mathematics', 'Physics', 'Linguistics',
                'Philosophy', 'Psychology', 'Religion')

Majors = c(67, 19, 22, 43, 53,
          58, 75, 36, 7, 25, 
          155, 6, 1, 3, 8, 
          154.5, 88, 99.5, 114, 42,
          63.5, 117, 16.5)

FTE = c(7.8, 4, 5.8, 3.7, 5.5, 
        9.5, 6.2, 3.4, 4, 4, 
        10, 5, 3, 3, 5, 
        9.8, 6.8, 9.55, 6, 3, 
        5.2, 7, 4)

major_data <- data.frame(Departments, Majors, FTE)
```

---

```{r}
major_data %>% ggplot(aes(x = Majors, y = FTE)) +
  geom_point() +
  ggtitle("Reed College Majors and FTE by Department")
```

---

```{r results="asis"}
# make a new data frame with only two columns to scatter plot 
keep <- c('Majors', 'FTE')
data2 <- major_data[keep]

# add names to new data frame as factor 
data2$pop.html.tooltip=major_data$Departments

# create interactive scatter plot using googleVis
Scatter1 <- gvisScatterChart(data2,                                                           
                            options=list(tooltip="{isHtml:'True'}",              # Define tooltip                            
                              legend="none", lineWidth=0, pointSize=5,                                                     
                              vAxis="{title:'Faculty (Total FTE)'}",             # y-axis label                
                              hAxis="{title:'Majors (delared and intended)'}",   # x-axis label                     
                              width=900, height=600))                            # plot dimensions  
print(Scatter1, 'chart') 
```

---

Left-click and drag to select an area of the chart to zoom on.  Right-click to zoom back out.

```{r results="asis"}
Scatter2 <- gvisScatterChart(data2,                                                           
                            options=list(
                              explorer="{actions: ['dragToZoom', 
                                          'rightClickToReset'],
                                           maxZoomIn:0.05}",
                              #chartArea="{width:'85%',height:'80%'}",
                              tooltip="{isHtml:'True'}",              
                              crosshair="{trigger:'both'}",                         
                              legend="none", lineWidth=0, pointSize=5,                                                     
                              vAxis="{title:'Faculty (Total FTE)'}",                        
                              hAxis="{title:'Majors (delared and intended)'}",                     
                              width=900, height=600))                                                                 
print(Scatter2, 'chart') 
```


---

### Alaskan departure delays in PNW

- The [`pnwflights14`](http://github.com/ismayc/pnwflights14) package provides information contains information about all flights that departed from SEA in Seattle and PDX in Portland, in 2014: 162,049 flights in total.

- We can use this data and the `dplyr` package to look at daily maximum departure delays throughout the year for Alaskan Airlines.

---

# Time series/line graphs

---

```{r message=FALSE, warning=FALSE}
data("flights", package = "pnwflights14")
alaskan <- flights %>% 
  filter(carrier %in% c("AS")) %>%
  mutate(date2014 = as.Date(paste0("2014/", month, "/", day))) %>%
  group_by(date2014) %>%
  summarize(max_dep_delay = max(dep_delay, na.rm=TRUE))
```

```{r}
alaskan %>% ggplot(aes(x = date2014, y = max_dep_delay)) +
  geom_line() +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %y") +
  xlab("Date") +
  ylab("Maximum Departure Delay")
```

---

```{r plotly=TRUE}
ggplotly()
```

---

## Plotting the time series using `dygraph` 

(Converting to time series format using `xts`)

```{r}
alaskan_ts <- xts(alaskan$max_dep_delay, alaskan$date2014)
colnames(alaskan_ts) <- "Max Departure Delay"
dygraph(alaskan_ts) %>% dyRangeSelector()
```

---

### Canadian and US population and geography

- Canada is an extremely large land mass (2nd largest country in the world), but is only the 37th largest country in terms of population

- The US ranks 4th highest in land mass and 3rd highest in population

- We can use data in the `maps` package to better visualize why these rankings exist

---

# Maps

---


```{r}
data(canada.cities, package = "maps")
canada_plot <- ggplot(canada.cities, aes(x = long, y = lat)) +
  coord_equal() +
  geom_point(aes(size=pop, text = paste0(name, ",",
    "Pop: ", prettyNum(pop, big.mark = ",", scientific = FALSE))), 
    colour = "red", alpha = 1/2) +
  borders(regions="canada")
canada_plot
```

---

```{r plotly=TRUE}
ggplotly(canada_plot)
```

---

```{r}
data(us.cities, package = "maps")
us_plot <- ggplot(us.cities, aes(x = long, y = lat)) +
  coord_equal() +
  geom_point(aes(size=pop, text = paste0(name, ",",
    "Pop: ", prettyNum(pop, big.mark = ",", scientific = FALSE))), 
    colour = "red", alpha = 1/2) +
  borders(regions="usa", xlim = c(-200, -60), ylim = c(20, 80))
us_plot
```

---

```{r plotly=TRUE}
ggplotly(us_plot)
```

---

# 3D objects

---

## New Zealand's highest volcano

```{r}
plot_ly(z = volcano, type = "surface")
```

---

# Interactive Data Tables

---

```{r}
datatable(iris, options = list(pageLength = 5))
```

---

## Another data table example

[RA Duty Scheduling](http://rpubs.com/cismay/ra_schedule_s2016)

---

## Other resources

[Plotting maps in R with `ggplot2`](https://github.com/dkahle/ggmap)

[HTML Widgets for R](http://www.htmlwidgets.org)

[Leaflet package for R](https://rstudio.github.io/leaflet/)

[GapMinder (now owned by Google)](http://www.google.com/publicdata/directory)

[Hans Rosling's TED talk - "The Best Stats You've Ever Seen"](https://www.ted.com/talks/hans_rosling_shows_the_best_stats_you_ve_ever_seen?language=en)

---

## What can I help you with?

> - Data analysis
> - Data wrangling/cleaning
> - Data visualization
> - Data tidying/manipulating
> - Reproducible research

---

## When am I available?

> - Email me at cismay@reed.edu or chester.ismay@reed.edu to schedule a time to meet if office hours don't work
> - Tentative Spring 2016 office (ETC 223) hours
>     - Mondays (10 AM to 11 AM)
>     - Tuesdays (2 PM to 3 PM)
>     - Wednesdays (1:30 PM to 2:30 PM)
> - Sometimes available for virtual office hours via Google Hangouts (email me for details)

---

# Thanks! 

<br>

<font size="7">cismay@reed.edu</font>

<br>

>- Code for slide creation on [my GitHub page](https://github.com/ismayc/Reed_WoRkshops/tree/master/Paideia_interactive_dataviz_2016)
>- Slides available [here](http://rpubs.com/cismay/paideia_2k16_idv)

---

```{r}
sessionInfo()
```

